Implementing Equational Constraints\\ in a Functional Language

### 2011-07-13 papers/wlp2011/paper.tex
WLP paper: section 4..7 refined

Due to this constructive definition, \ccode{=:=}
can be considered as a binary function defined by the
following rules (here we show only the rules for the Boolean
and list types, where \code{Success} denotes the only constructor
of the type \code{Success} of constraints):
\begin{curry}
  True   =:= True    =  Success
  False  =:= False   =  Success
  []     =:= []      =  Success
  (x:xs) =:= (y:ys)  =  x =:= y & xs =:= ys

@@

\end{haskell}
%
Although this is a correct implementation of equational constraints,
it might lead to an unnecessarily large search space
when it is applied to generators representing logic variables.
For instance, consider the following generator for Boolean lists:
\begin{curry}
  aBoolList = [] ? (aBool : aBoolList)
\end{curry}
This is translated into Haskell as follows:
\begin{haskell}
  aBoolList :: IDSupply -> [Bool]
  aBoolList s = Choice (thisID s) [] (aBool (leftSupply s)
                                      : aBoolList (rightSupply s))
\end{haskell}
Now consider the equational constraint \ccode{x =:= [True]}.
If the logic variable \code{x} is replaced by \code{aBoolList},
the translated expression \ccode{aBoolList s =:= [True]}
creates a search space when evaluating its first argument,
although there is no search required since there is only one possible
binding for \code{x}.
Furthermore and even worse, there is an infinite search space
when unifying two logic variables. For instance,
the expression \ccode{xs =:= ys \& xs++ys =:= [True]}
results in an infinite search space when the logic variables
\code{xs} and \code{ys} are replaced by generators.

To avoid these problems, we have to implement the idea
of the well-known unification principle \cite{Robinson65}.
Instead of enumerating all values for logic variables
occurring in an equational constraint,
we \emph{bind} them to another variable or term.
Since we compile into a purely functional language,
the binding cannot be performed by some side effect.
Instead, we add binding constraints to the computed results.

To implement unification, we have to distinguish free variables
from ``standard choices'' (introduced by the operation \ccode{?})
in the target code. For this purpose, we refine the definition
of the type \code{ID} as follows:\footnote{For the sake
of simplicity, we consider in the following the
implementation of \code{IDSupply} as unbounded integers.}
\begin{haskell}
  data ID = ChoiceID Integer | FreeID Integer
\end{haskell}
The new constructor \code{FreeID} marks a choice corresponding to
a free variable, e.g., the generator for Boolean variables is
redefined as
\begin{haskell}
  aBool s = Choice (FreeID (thisID s)) True False
\end{haskell}
%
If an operation is applied to a free variable and requires its value,
the free variable is transformed into a standard choice.
For this purpose, we define a simple operation to perform
this transformation:
\begin{haskell}
  narrow :: ID -> ID
  narrow (FreeID i) = ChoiceID i
  narrow x          = x
\end{haskell}
Furthermore, we use this operation in narrowing steps,
i.e., in all rules operating on \code{Choice} constructors.
For instance, in the implementation of the operation \code{not}
we replace the rule
\begin{haskell}
  not (Choice i x1 x2) s = Choice i (not x1 s) (not x2 s)
\end{haskell}
by the rule
\begin{haskell}
  not (Choice i x1 x2) s = Choice (narrow i) (not x1 s) (not x2 s)
\end{haskell}
With this small change, the run-time system can distinguish
free variables (choices with an identifier of the form \code{(FreeID $\ldots$)})
from other values. This is quite useful at the top-level
to print values containing logic variables in a specific form
rather than enumerating all values, similarly to logic programming systems.
For instance, KiCS2 evaluates the application of \code{head}
to an unknown list as follows:
\begin{haskell}
  Prelude> head xs where xs free
  {xs = (_x2:_x3)} _x2
\end{haskell}
Here, free variables are marked by the prefix \code{\us{}x}.

As mentioned above, the consideration of free variables
is relevant in equational constraints where \emph{binding constraints} are
generated. For this purpose, we introduce a type to represent
a binding constraint as a pair of a choice identifier
and a decision for this choice:
\begin{haskell}
  data Constraint = ID :=: Decision
\end{haskell}
Furthermore, we extend each data type by the possibility to
add such a constraint:
\begin{haskell}
  data Bool   = $\ldots$ | Guard Constraint Bool
  data List a = $\ldots$ | Guard Constraint (List a)
\end{haskell}
Thus, \code{(Guard $c$ $v$)} represents a \emph{constrained value},
i.e., the value $v$ is only valid if the constraint $c$ is
consistent with the decisions previously made during search.
The binding constraints are created by the
equational constraint operation \ccode{=:=}: if a free variable
should be bound to a constructor, we make the same
decision as it would be done in the successful branch
of the generator. In case of Boolean values,
this can be implemented by the following additional rules
for \ccode{=:=}:
\begin{haskell}
  Choice (FreeID i) _ _ =:= True   =  Guard (i :=: ChooseLeft ) Success
  Choice (FreeID i) _ _ =:= False  =  Guard (i :=: ChooseRight) Success
\end{haskell}
Hence, the binding of a variable to some known value
is implemented as a binding constraint for the choice identifier
for this variable. However, if we want to bind a variable
to another variable, we cannot store a concrete decision.
Instead, we store the information that the decisions for
both variables, when they are made to extract values,
must be identical. For this purpose, we extend the \code{Decision}
type to cover this information:
\begin{haskell}
  data Decision = $\ldots$ | BindTo ID
\end{haskell}
Furthermore, we add the rule that an equational constraint
between two variables yields a binding for these variables:
\begin{haskell}
  Choice (FreeID i) _ _ =:= Choice (FreeID j) _ _
    =  Guard (i :=: BindTo j) Success
\end{haskell}
%
The consistency of constraints is checked when values are extracted
from a choice structure, e.g., by the operation \code{printValsDFS}.
For this purpose, we extend the definition of the corresponding search
operations by calling a constraint solver for the constraints.
For instance, the definition of \code{printValsDFS} is extended by
a rule handling constrained values:
%
\begin{haskell}
  $\ldots$
  printValsDFS (Guard c x) = do consistent <- add c
                                if consistent then do printValsDFS x
                                                      remove c
                                              else return ()
  $\ldots$
\end{haskell}
The operation \code{add} checks the consistency of the constraint \code{c}
with the decisions made so far and, in case of consistency,
stores the decision made by the constraint.
In this case, the constrained value is evaluated
before the constraint is removed to allow backtracking.
Furthermore, the operations \code{lookupDecision} and \code{setDecision}
are extended to deal with bindings between two variables,
i.e., they follow variable chains in case of \code{BindTo} constructors.

Due to the extensions described in this section,
we obtain an implementation of a unification procedure
that integrates the demand-driven evaluation of operations
occurring in the arguments.
Benchmarks evaluating the efficiency of this approach are
shown in Section~\ref{sec:Benchmarks}.

@@

A well-known disadvantage of equational constraints
is the fact that \ccode{=:=} is interpreted as strict equality.
Thus, if one uses equational constraints to express
requirements on arguments, the resulting operations might be too strict.
For instance, the equational constraint in the condition
defining \code{last} (see Section~\ref{ex:last})
requires that \code{ys++[e]} as well as \code{xs}
must be reducible to identical values so that the input list \code{xs}
is completely evaluated.
Hence, if \code{failed} denotes an operation whose evaluation fails,
the evaluation of \code{last [failed,True]} has no result.
On the other hand, the evaluation of \code{last' [failed,True]} yields
the value \code{True}, i.e., the definition of \code{last'} is less strict.
This is due to the fact that a functional pattern like
\code{(xs++[e])} abbreviates all values to which it can be evaluated
(by narrowing), like \code{[e]}, \code{[x1,e]}, \code{[x1,x2,e]} etc.
Conceptually, the rule defining \code{last'}
abbreviates the following (infinite) set of rules:
\begin{curry}
  last' [e] = e
  last' [x1,e] = e
  last' [x1,x2,e] = e
  $\ldots$
\end{curry}
As a further example,
consider an operation that returns the first duplicated element in
a list.
Using equational constraints, we can define it as follows:
\begin{curry}
  fstDup xs | xs =:= ys++[e]++zs & elem e ys =:= True & nub ys =:= ys
            = e    where ys, zs, e free
\end{curry}
The first equational constraint is used to split the input list \code{xs}
into three sublists.
The last constraint ensures that the first sublist
\code{ys} does not contain duplicated elements
(the library operation \code{nub} removes all duplicates from a list)
and the second constraint ensures that the first element after \code{ys}
occurs in \code{ys}.
Although this implementation is concise, it cannot be applied
to infinite lists due to the strict interpretation of \ccode{=:=}.
This is not the case if we define this operation by a functional pattern:
\begin{curry}
  fstDup' (ys++[e]++zs) | elem e ys =:= True & nub ys =:= ys
                        = e
\end{curry}
%
Functional patterns are quite useful in situations
where one wants to select parts of some structure
at arbitrary positions. For instance,
functional patterns have been applied to support
the high-level programming of querying or transforming
XML data \cite{Hanus11ICLP}.

Obviously, one cannot implement functional patterns by
a transformation into an infinite set of rules. Instead, they are
implemented by a specific \emph{lazy unification} procedure \ccode{=:<=}
\cite{AntoyHanus05LOPSTR}.
For instance, the definition of \code{last'} is transformed into
\begin{curry}
  last' ys | (xs++[e]) =:<= ys  = e   where xs, e free
\end{curry}
The behavior of \ccode{=:<=} is similar to \ccode{=:=}
except for the case that a variable in the left argument
should be bound to some expression: instead of evaluating
the expression to some value and binding the variable to the value,
the variable is bound to the \emph{unevaluated} expression
(see \cite{AntoyHanus05LOPSTR} for more details).
Due to this slight change, failures or infinite structures
in actual arguments do not cause problems in the matching
of functional patterns.

The following discussion shows the requirements on an 
implementation of functional patterns.
The general structure
is quite similar to equational constraints
with the exception that variables could be also bound
to unevaluated expressions. Only if such variables are later
accessed, the expressions they are bound to are evaluated.
This can be achieved by adding a further alternative
to the type of decisions:
\begin{haskell}
  data Decision = $\ldots$ | LazyBind Constraint
\end{haskell}
%
The implementation of the lazy unification operation \ccode{=:<=}
is almost identical to the strict unification operation \ccode{=:=}
as shown in Section~\ref{sec:Unification}.
The only difference is in the rules where a free variable occurs
in the left argument. All these rules are replaced by the single
rule
\begin{haskell}
  Choice (FreeID i) _ _ =:<= x
    = Guard (i :=: LazyBind (lazyBind i x)) Success
\end{haskell}
where the auxiliary operation \code{lazyBind}
implements the demand-driven evaluation of the right argument \code{x}:
\begin{haskell}
  lazyBind :: ID -> a -> Constraint
  lazyBind i True  = i :=: ChooseLeft
  lazyBind i False = i :=: ChooseRight
\end{haskell}
The use of the additional \code{LazyBind} constructor allows the argument 
\code{x} to be stored in a binding constraint without evaluation 
(due to the lazy evaluation strategy of the
target language Haskell).
However, it is evaluated by \code{lazyBind} when its binding
is required by another part of the computation.
Similarly to equational constraints,
lazy bindings are processed by a solver when values are extracted.
In particular, if a variable has more than one lazy binding constraint
(which is possible if a functional pattern evaluates to a non-linear term),
the corresponding expressions are evaluated and unified
according to the semantics of functional patterns
\cite{AntoyHanus05LOPSTR}.

In order to demonstrate the operational behavior of our implementation,
we sketch the evaluation of the lazy unification constraint
\code{xs++[e] =:<= [failed,True]} that occurs when the expression
\code{last' [failed,True]} is evaluated (here we omit failed branches
and some other details; note that logic variables are replaced
by generators):
\begin{haskell}
  aBoolList 2 ++ [aBool 3] =:<= [failed,True]
  $\leadsto$ [aBool 4,aBool 3] =:<= [failed,True]
  $\leadsto$ aBool 4 =:<= failed & aBool 3 =:<= True & [] =:<= []
  $\leadsto$ Guard (4 :=: LazyBind (lazyBind 4 failed))
           (Guard (3 :=: LazyBind (lazyBind 3 True)) Success)
\end{haskell}
If the binding of identifier \code{3} is later required,
\code{(lazyBind 3 True)} is evaluated to \code{[3 :=: ChooseLeft]}
which corresponds to the value \code{True} of the generator \code{(aBool 3)}.
Note that the variable with identifier \code{4} does not occur
anywhere else so that the binding \code{(lazyBind 4 failed)}
will never be evaluated, as intended.

@@

In this section we evaluate our implementation by some benchmarks.
They were executed on a Linux machine running Debian 5.0.7 with
an Intel Core 2 Duo (3.0GHz) processor.
KiCS2 has been executed with
the Glasgow Haskell Compiler (GHC 7.0.4, option -O2) and uses
an efficient \code{IDSupply} implementation that makes use of \code{IORef}s.
For a comparison with other mature implementations of Curry,
we considered PAKCS \cite{Hanus10PAKCS}
(Version 1.9.2, based on a SICStus-Prolog 4.1.2)
and MCC \cite{Lux99FLOPS} (Version 0.9.10). The timings were performed with the
time command measuring the execution time (in seconds) of a compiled
executable for each benchmark as a mean of three runs.
The programs used for the benchmarks,
partially taken from \cite{AntoyHanus05LOPSTR},
are \code{last} (compute the last element of a list),\footnote{%
\ccode{inc x n} is a naive addition that \code{n} times
increases its argument \code{x} by 1.}
\code{simplify} (simplify a symbolic arithmetic expression),
\code{varInExp} (non-deterministically return a variable occuring
                  in a symbolic arithmetic expression),
\code{half} (compute the half of a Peano number using logic variables),
\code{palindrome} (check whether a list is a palindrome),
\code{horseman} (solving an equation relating heads and feet of horses and men
based on Peano numbers),
and
\code{grep} (string matching based on a non-deterministic
specification of regular expressions \cite{AntoyHanus10CACM}).

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|r|}
\hline
Expression                              & \code{==~}& \code{=:=}& \code{=:<=} \\
\hline
\code{last (map (inc 0) [1..100000])}   & 134.23    & 0.83      & 0.26        \\
\code{simplify}                         & 8.50      & 7.15      & 7.34        \\
\code{varInExp}                         & 11.72     & 1.50      & 1.40        \\
% \code{varInExp2}                        & 423.01    & 10.50     & 1.04        \\
% \code{varInExp3}                        & 399.68    & 16.70     & 11.28       \\
\code{fromPeano (half (toPeano 10000))} & 32.92     & 6.35      & 11.88       \\
\code{palindrome}                       & 28.59     & 15.58     & 23.21       \\
\code{horseman}                         & 3.58      & 3.13      & n/a         \\
\code{grep}                             & 1.06      & 0.19      & n/a         \\
\hline
\end{tabular}
\caption{Benchmarks: comparing different equation solving methods}
 \label{fig:search-strategies}
\end{figure}

In Section~\ref{sec:Unification} we mentioned that equational
constraints could also be solved by generators without variable bindings,
but this technique might increase the search space due to the
possibly superfluous generation of all values.
To show the beneficial effects of our implementation
of equational constraints with variable bindings,
we compare in Fig.~\ref{fig:search-strategies}
the results of using equational constraints (\code{=:=})
to the results
where the Boolean equality operator (\code{==}) is used (which
does not perform bindings but enumerate all values).
As expected, the creation and traversal of a large search space introduced
by \code{==} is in most cases much slower than our presented approach
with variable bindings.
In addition, the example \code{last} shows that the lazy 
unification operator (\code{=:<=}) improves performance when unifying an
expression which has to be evaluted only partially. Using strict unification,
all elements of the list are (unnecessarily) evaluated.

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|r|}
\hline
Expression                              & KiCS2 & PAKCS  & MCC  \\
\hline
\code{last (map (inc 0) [1..100000])}   & 0.83  & 4.25   & 0.14 \\
\code{simplify}                         & 7.15  & 0.15   & 0.00 \\
\code{varInExp}                         & 1.50  & 3.89   & 0.30 \\
% \code{varInExp2}                        & 10.50 & 3.48   & stckOflw \\
% \code{varInExp3}                        & 16.70 & 19.22  & 1.18 \\
\code{fromPeano (half (toPeano 10000))} & 6.35  & 108.88 & 3.22 \\
\code{palindrome}                       & 15.58 & 32.56  & 1.07 \\
\code{horseman}                         & 3.13  & 8.70   & 0.42 \\
\code{grep}                             & 0.19  & 2.88   & 0.14 \\
\hline
\end{tabular}
\caption{Benchmarks: equational constraints in different Curry implementations}
 \label{fig:unification}
\end{figure}

In contrast to the Curry implementations PAKCS and MCC, our approach is based
on an explicit representation of the search space instead of
backtracking and manipulating
a global state containing bindings for logic variables.
Nevertheless, the benchmarks in Fig.~\ref{fig:unification}
show that it can compete with or even outperform these implementations.
In these benchmarks only equational constraints are used.
The results show that the implementation of unification
of MCC performs best. However, in most cases
our implementation outperforms the Prolog-based PAKCS implementation,
except for some examples where further investigations
and optimizations are necessary.

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|}
\hline
Expression                              & KiCS2 & PAKCS    \\
\hline
\code{last (map (inc 0) [1..100000])}   & 0.26  & 3.97     \\
\code{simplify}                         & 7.34  & 0.27     \\
\code{varInExp}                         & 1.40  & 8.88     \\
% \code{varInExp2}                         & 1.04  & 0.08    \\
% \code{varInExp3}                         & 11.28 & 16.62   \\
\code{fromPeano (half (toPeano 10000))} & 11.88 & $\infty$ \\
\code{palindrome}                       & 23.21 & $\infty$ \\
\hline
\end{tabular}
\caption{Benchmarks: comparing different implementations of functional patterns}
 \label{fig:functional-patterns}
\end{figure}

@@

For future work it might be interesting to add
further constraint structures, like real arithmetic
or finite domain constraints, to our implementation.
This might be possible by extending the kinds of constraints
of our implementation and solving them by
functional approaches like \cite{SchrijversStuckeyWadler09}.

>>>>

Due to this constructive definition, \ccode{=:=}
can be considered as a binary function defined by the
following rules (we only present the rules for the Boolean
and list types, where \code{Success} denotes the only constructor
of the type \code{Success} of constraints):
\begin{curry}
  True   =:= True    =  Success
  False  =:= False   =  Success

  []     =:= []      =  Success
  (x:xs) =:= (y:ys)  =  x =:= y & xs =:= ys

@@

\end{haskell}
%
Although this is a correct implementation of equational constraints,
it might lead to an unnecessarily large search space
when it is applied to generators representing logic variables.
For instance, consider the following generator for Boolean lists:
\begin{curry}
  aBoolList = [] ? (aBool : aBoolList)
\end{curry}
This is translated into Haskell as follows:
\begin{haskell}
  aBoolList :: IDSupply -> [Bool]
  aBoolList s = Choice (thisID s) [] (aBool (leftSupply s)
                                      : aBoolList (rightSupply s))
\end{haskell}
Now consider the equational constraint \ccode{x =:= [True]}.
If the logic variable \code{x} is replaced by \code{aBoolList},
the translated expression \ccode{aBoolList s =:= [True]}
creates a search space when evaluating its first argument,
although there is no search required since there is only one
binding for \code{x} satisfying the constraint.
Furthermore and even worse, unifying two logic variables
introduces an infinite search space. For instance,
the expression \ccode{xs =:= ys \& xs++ys =:= [True]}
results in an infinite search space when the logic variables
\code{xs} and \code{ys} are replaced by generators.

To avoid these problems, we have to implement the idea
of the well-known unification principle \cite{Robinson65}.
Instead of enumerating all values for logic variables
occurring in an equational constraint,
we \emph{bind} the variables to another variable or term.
Since we compile into a purely functional language,
the binding cannot be performed by some side effect.
Instead, we add binding constraints to the computed results to be 
processed during search.

To implement unification, we have to distinguish free variables
from ``standard choices'' (introduced by overlapping rules)
in the target code. For this purpose, we refine the definition
of the type \code{ID} as follows:\footnote{For the sake
of simplicity, in the following, we consider the
implementation of \code{IDSupply} to be unbounded integers.}
\begin{haskell}
  data ID = ChoiceID Integer | FreeID Integer
\end{haskell}
The new constructor \code{FreeID} identifies a choice corresponding to
a free variable, e.g., the generator for Boolean variables is
redefined as
\begin{haskell}
  aBool s = Choice (FreeID (thisID s)) True False
\end{haskell}
%
If an operation is applied to a free variable and requires its value,
the free variable is transformed into a standard choice.
For this purpose, we define a simple operation to perform
this transformation:
\begin{haskell}
  narrow :: ID -> ID
  narrow (FreeID i) = ChoiceID i
  narrow x          = x
\end{haskell}
Furthermore, we use this operation in narrowing steps,
i.e., in all rules operating on \code{Choice} constructors.
For instance, in the implementation of the operation \code{not}
we replace the rule
\begin{haskell}
  not (Choice i x1 x2) s = Choice i (not x1 s) (not x2 s)
\end{haskell}
by the rule
\begin{haskell}
  not (Choice i x1 x2) s = Choice (narrow i) (not x1 s) (not x2 s)
\end{haskell}

As mentioned above, the consideration of free variables
is relevant in equational constraints where \emph{binding constraints} are
generated. For this purpose, we introduce a type to represent
a binding constraint as a pair of a choice identifier
and a decision for this identifier:
\begin{haskell}
  data Constraint = ID :=: Decision
\end{haskell}
Furthermore, we extend each data type by the possibility to
add constraints:
\begin{haskell}
  data Bool   = $\ldots$ | Guard [Constraint] Bool
  data List a = $\ldots$ | Guard [Constraint] (List a)
\end{haskell}
Thus, \code{(Guard $cs$ $v$)} represents a \emph{constrained value},
i.e., the value $v$ is only valid if the constraints $cs$ are
consistent with the decisions previously made during search.
These binding constraints are created by the
equational constraint operation \ccode{=:=}: if a free variable
should be bound to a constructor, we make the same
decisions as it would be done in the successful branch
of the generator. In case of Boolean values,
this can be implemented by the following additional rules
for \ccode{=:=}:
\begin{haskell}
  Choice (FreeID i) _ _ =:= True   =  Guard [i :=: ChooseLeft ] Success
  Choice (FreeID i) _ _ =:= False  =  Guard [i :=: ChooseRight] Success
\end{haskell}
Hence, the binding of a variable to some known value
is implemented as a binding constraint for the choice identifier
for this variable. However, if we want to bind a variable
to another variable, we cannot store a concrete decision.
Instead, we store the information that the decisions for
both variables, when they are made to extract values,
must be identical. For this purpose, we extend the \code{Decision}
type to cover this information:
\begin{haskell}
  data Decision = $\ldots$ | BindTo ID
\end{haskell}
Furthermore, we add the rule that an equational constraint
between two variables yields a binding for these variables:
\begin{haskell}
  Choice (FreeID i) _ _ =:= Choice (FreeID j) _ _
    =  Guard (i :=: BindTo j) Success
\end{haskell}
%
The consistency of constraints is checked when values are extracted
from a choice structure, e.g., by the operation \code{printValsDFS}.
For this purpose, we extend the definition of the corresponding search
operations by calling a constraint solver for the constraints.
For instance, the definition of \code{printValsDFS} is extended by
a rule handling constrained values:
%
\begin{haskell}
  $\ldots$
  printValsDFS (Guard cs x) = do consistent <- add c
                                 if consistent then do printValsDFS x
                                                       remove c
                                               else return ()
  $\ldots$
\end{haskell}
The operation \code{add} checks the consistency of the constraint \code{c}
with the decisions made so far and, in case of consistency,
stores the decision made by the constraint.
In this case, the constrained value is evaluated
before the constraint is removed to allow backtracking.
Furthermore, the operations \code{lookupDecision} and \code{setDecision}
are extended to deal with bindings between two variables,
i.e., they follow variable chains in case of \code{BindTo} constructors.

Finally, with the ability to distinguish free variables 
(choices with an identifier of the form \code{(FreeID $\ldots$)})
from other values during search, values containing logic variables 
can also be printed in a specific form rather than enumerating all values, 
like in logic programming systems. For instance, KiCS2 evaluates the 
application of \code{head} to an unknown list as follows:
\begin{haskell}
  Prelude> head xs where xs free
  {xs = (_x2:_x3)} _x2
\end{haskell}
Here, free variables are marked by the prefix \code{\us{}x}.

% Due to the extensions described in this section,
% we obtain an implementation of a unification procedure
% that integrates the demand-driven evaluation of operations
% occurring in the arguments.
% Benchmarks evaluating the efficiency of this approach are
% shown in Section~\ref{sec:Benchmarks}.

@@

A well-known disadvantage of equational constraints
is the fact that \ccode{=:=} is interpreted as strict equality.
Thus, if one uses equational constraints to express
requirements on arguments, the resulting operations might be too strict.
For instance, the equational constraint in the condition
defining \code{last} (see Section~\ref{ex:last})
requires that \code{ys++[e]} as well as \code{xs}
must be reducible to unifiable terms so that in consequence
the input list \code{xs} is completely evaluated.
Hence, if \code{failed} denotes an operation whose evaluation fails,
the evaluation of \code{last [failed,True]} has no result.
On the other hand, the evaluation of \code{last' [failed,True]} yields
the value \code{True}, i.e., the definition of \code{last'} is less strict.

As another example for the advantage of the reduced strictness,
consider an operation that returns the first duplicate element in
a list.
Using equational constraints, we can define it as follows:
\begin{curry}
  fstDup xs | xs =:= ys++[e]++zs & elem e ys =:= True & nub ys =:= ys
            = e    where ys, zs, e free
\end{curry}
The first equational constraint is used to split the input list \code{xs}
into three sublists.
The last constraint ensures that the first sublist
\code{ys} does not contain duplicated elements
(the library operation \code{nub} removes all duplicates from a list)
and the second constraint ensures that the first element after \code{ys}
occurs in \code{ys}.
Although this implementation is concise, it cannot be applied
to infinite lists due to the strict interpretation of \ccode{=:=}.
This is not the case if we define this operation by a functional pattern:
\begin{curry}
  fstDup' (ys++[e]++zs) | elem e ys =:= True & nub ys =:= ys
                        = e
\end{curry}
%
Because of the reduced strictness, the logic variable \code{zs} (matching
the tail list after the first duplicate) is never evaluated.
This is due to the fact that a functional pattern like
\code{(xs++[e])} abbreviates all values to which it can be evaluated
(by narrowing), like \code{[e]}, \code{[x1,e]}, \code{[x1,x2,e]} etc.
Conceptually, the rule defining \code{last'}
abbreviates the following (infinite) set of rules:
\begin{curry}
  last' [e] = e
  last' [x1,e] = e
  last' [x1,x2,e] = e
  $\ldots$
\end{curry}

Obviously, one cannot implement functional patterns by
a transformation into an infinite set of rules. Instead, they are
implemented by a specific \emph{lazy unification} procedure \ccode{=:<=}
\cite{AntoyHanus05LOPSTR}.
For instance, the definition of \code{last'} is transformed into
\begin{curry}
  last' ys | (xs++[e]) =:<= ys  = e   where xs, e free
\end{curry}
The behavior of \ccode{=:<=} is similar to \ccode{=:=},
except for the case that a variable in the left argument
should be bound to some expression: instead of evaluating
the expression to some value and binding the variable to the value,
the variable is bound to the \emph{unevaluated} expression
(see \cite{AntoyHanus05LOPSTR} for more details).
Due to this slight change, failures or infinite structures
in actual arguments do not cause problems in the matching
of functional patterns.

The general structure of the implementation of functional patterns in KiCS2
is quite similar to that of equational constraints,
with the exception that variables could be also bound
to unevaluated expressions. 
Only if such variables are later
accessed, the expressions they are bound to are evaluated.
This can be achieved by adding a further alternative
to the type of decisions:
\begin{haskell}
  data Decision = $\ldots$ | LazyBind [Constraint]
\end{haskell}
%
The implementation of the lazy unification operation \ccode{=:<=}
is almost identical to the strict unification operation \ccode{=:=}
as shown in Section~\ref{sec:Unification}.
The only difference is in the rules where a free variable occurs
in the left argument. All these rules are replaced by the single
rule
\begin{haskell}
  Choice (FreeID i) _ _ =:<= x
    = Guard [i :=: LazyBind (lazyBind i x)] Success
\end{haskell}
where the auxiliary operation \code{lazyBind}
implements the demand-driven evaluation of the right argument \code{x}:
\begin{haskell}
  lazyBind :: ID -> a -> [Constraint]
  lazyBind i True  = i :=: [ChooseLeft]
  lazyBind i False = i :=: [ChooseRight]
\end{haskell}
The use of the additional \code{LazyBind} constructor allows the argument 
\code{x} to be stored in a binding constraint without evaluation 
(due to the lazy evaluation strategy of the
target language Haskell).
However, it is evaluated by \code{lazyBind} when its binding
is required by another part of the computation.
Similarly to equational constraints,
lazy bindings are processed by a solver when values are extracted.
In particular, if a variable has more than one lazy binding constraint
(which is possible if a functional pattern evaluates to a non-linear term),
the corresponding expressions are evaluated and unified
according to the semantics of functional patterns
\cite{AntoyHanus05LOPSTR}.

In order to demonstrate the operational behavior of our implementation,
we sketch the evaluation of the lazy unification constraint
\code{xs++[e] =:<= [failed,True]} that occurs when the expression
\code{last' [failed,True]} is evaluated (we omit failed branches
and some other details; note that logic variables are replaced
by generators):
\begin{haskell}
     aBoolList 2 ++ [aBool 3] =:<= [failed, True]
  $\leadsto$ [aBool 4, aBool 3] =:<= [failed, True]
  $\leadsto$ aBool 4 =:<= failed & aBool 3 =:<= True & [] =:<= []
  $\leadsto$ Guard [ 4 :=: LazyBind (lazyBind 4 failed)
           , 3 :=: LazyBind (lazyBind 3 True)] Success
\end{haskell}
If the value of the expression \code{last' [failed,True]} is later required,
the value of the variable \code{e} (with the identifier \code{3}) is in turn
required.
Thus, \code{(lazyBind 3 True)} is evaluated to \code{[3 :=: ChooseLeft]}
which corresponds to the value \code{True} of the generator \code{(aBool 3)}.
Note that the variable with identifier \code{4} does not occur
anywhere else, so that the binding \code{(lazyBind 4 failed)}
will never be evaluated, as intended.

@@

In this section we evaluate our implementation of
equational constraints and functional patterns by some benchmarks.
The benchmarks were executed on a Linux machine running Debian 5.0.7 with
an Intel Core 2 Duo (3.0GHz) processor.
KiCS2 has been used with
the Glasgow Haskell Compiler (GHC 7.0.4, option -O2) as its backend and
an efficient \code{IDSupply} implementation that makes use of \code{IORef}s.
For a comparison with other mature implementations of Curry,
we considered PAKCS \cite{Hanus10PAKCS}
(version 1.9.2, based on a SICStus-Prolog 4.1.2)
and MCC \cite{Lux99FLOPS} (version 0.9.10). The timings were performed with the
time command measuring the execution time (in seconds) of a compiled
executable for each benchmark as a mean of three runs.
The programs used for the benchmarks,
partially taken from \cite{AntoyHanus05LOPSTR},
are \code{last} (compute the last element of a list),\footnote{%
\ccode{inc x n} is a naive addition that \code{n} times
increases its argument \code{x} by 1.}
\code{simplify} (simplify a symbolic arithmetic expression),
\code{varInExp} (non-deterministically return a variable occuring
                  in a symbolic arithmetic expression),
\code{half} (compute the half of a Peano number using logic variables),
\code{palindrome} (check whether a list is a palindrome),
\code{horseman} (solving an equation relating heads and feet of horses and men
based on Peano numbers),
and
\code{grep} (string matching based on a non-deterministic
specification of regular expressions \cite{AntoyHanus10CACM}).

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|r|}
\hline
Expression                              & \code{==~}& \code{=:=}& \code{=:<=} \\
\hline
\code{last (map (inc 0) [1..100000])}   & 134.23    & 0.83      & 0.26        \\
\code{simplify}                         & 8.50      & 7.15      & 7.34        \\
\code{varInExp}                         & 11.72     & 1.50      & 1.40        \\
% \code{varInExp2}                        & 423.01    & 10.50     & 1.04        \\
% \code{varInExp3}                        & 399.68    & 16.70     & 11.28       \\
\code{fromPeano (half (toPeano 10000))} & 32.92     & 6.35      & 11.88       \\
\code{palindrome}                       & 28.59     & 15.58     & 23.21       \\
\code{horseman}                         & 3.58      & 3.13      & n/a         \\
\code{grep}                             & 1.06      & 0.19      & n/a         \\
\hline
\end{tabular}
\caption{Benchmarks: comparing different representations for equations}
 \label{fig:search-strategies}
\end{figure}

In Section~\ref{sec:Unification} we mentioned that equational
constraints could also be solved by generators without variable bindings,
but this technique might increase the search space due to the
possibly superfluous generation of all values.
To show the beneficial effects of our implementation
of equational constraints with variable bindings,
in Fig.~\ref{fig:search-strategies} we compare
the results of using equational constraints (\code{=:=})
to the results
where the Boolean equality operator (\code{==}) is used (which
does not perform bindings but enumerate all values).
As expected, in most cases the creation and traversal of a large
search space introduced by \code{==} is much slower than our
presented approach with variable bindings.
In addition, the example \code{last} shows that the lazy 
unification operator (\ccode{=:<=}) improves the performance when unifying an
expression which has to be evaluted only partially. Using strict unification,
all elements of the list are (unnecessarily) evaluated.

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|r|}
\hline
Expression                              & KiCS2 & PAKCS  & MCC  \\
\hline
\code{last (map (inc 0) [1..100000])}   & 0.83  & 4.25   & 0.14 \\
\code{simplify}                         & 7.15  & 0.15   & 0.00 \\
\code{varInExp}                         & 1.50  & 3.89   & 0.30 \\
% \code{varInExp2}                        & 10.50 & 3.48   & stckOflw \\
% \code{varInExp3}                        & 16.70 & 19.22  & 1.18 \\
\code{fromPeano (half (toPeano 10000))} & 6.35  & 108.88 & 3.22 \\
\code{palindrome}                       & 15.58 & 32.56  & 1.07 \\
\code{horseman}                         & 3.13  & 8.70   & 0.42 \\
\code{grep}                             & 0.19  & 2.88   & 0.14 \\
\hline
\end{tabular}
\caption{Benchmarks: strict unification in different Curry implementations}
 \label{fig:unification}
\end{figure}

In contrast to the Curry implementations PAKCS and MCC, our approach 
of strict unification is based
on an explicit representation of the search space instead of
backtracking and manipulating
a global state containing bindings for logic variables.
Nevertheless, the benchmarks in Fig.~\ref{fig:unification},
using equational constraints only,
show that it can compete with or even outperform the other implementations.
The results show that the implementation of unification
of MCC performs best. However, in most cases our implementation 
outperforms the Prolog-based PAKCS implementation, 
except for some examples where further investigations
and optimizations are necessary.

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|}
\hline
Expression                              & KiCS2 & PAKCS    \\
\hline
\code{last (map (inc 0) [1..100000])}   & 0.26  & 3.97     \\
\code{simplify}                         & 7.34  & 0.27     \\
\code{varInExp}                         & 1.40  & 8.88     \\
% \code{varInExp2}                         & 1.04  & 0.08    \\
% \code{varInExp3}                         & 11.28 & 16.62   \\
\code{fromPeano (half (toPeano 10000))} & 11.88 & $\infty$ \\
\code{palindrome}                       & 23.21 & $\infty$ \\
\hline
\end{tabular}
\caption{Benchmarks: functional patterns in different Curry implementations}
 \label{fig:functional-patterns}
\end{figure}

@@

For future work it might be interesting to add
further constraint structures to our implementation, like real arithmetic
or finite domain constraints.
This might be possible by extending the kinds of constraints
of our implementation and solving them by
functional approaches like \cite{SchrijversStuckeyWadler09}.
